An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-05-09 Project_0/manuscript/manuscript.Rmd
cut out unused text

<!--
Similar to the AIC results, bias results only condition on whether the estimating model completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibility of the hessian because the optimized model can provide reliable estimation of assessment model output whether or not the model was able to estimate non-zero random effects. In practice, the model would be reconfigured to remove unnecessary process errors and produce otherwise equivalent parameter estimates.
-->

@@

We calculated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM [@mohn99]. We fit 7 peels for each EM and calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias. <!--Similar to the other results, retrospective results only condition on whether all of the peels of a given estimating model completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibility of the hessian because the optimized model can provide reliable estimation of SSB, fishing mortality rates, and recruitment whether or not the model was able to estimate non-zero random effects. -->

@@

<!--
Finally, although the EMs estimated observation variance parameters for the age composition observations, they were provided the correct precision of the aggregate indices and catch which

The R+q OMs only have temporal variation in catchability for one index so the other index is able to provide a consistent signal of trend. Incorrect assumptions about the mean selectivity model could also produce

Lack of retrospective patterns for SSB coupled with biased terminal SSB estimates would indicate that the scale across the time series is biased and
-->

@@

<!--
All together what do results mean? E.g., when do you get good convergence and retrospective patterns and high/none bias? When do you get good convergence and no retros and no/big bias?

For R+S OMs with low observation error, temporal contrast in fishing pressure, and large variation in survival process errors (3rd row), some EMs with the incorrect process error assumptions, estimate median M to be 0, and SSB is also underestimated, but Mohn's $\rho$ for SSB is better than when M is known. Fortunately, AIC would not choose the wrong EM process error type in these situations.

When R+S is preferred AMONG THE TYPES CONSIDERED HERE, it probably means its the correct process error .
-->

<!--
## Summary {-}

We conducted a simulation study where we simulated process errors and observations for 72 operating models with alternative assumptions about fishing history, degree of uncertainty in index and age composition observations, type (recruitment, survival, fishery selectivity, catchability and natural mortality), degree of variation, and correlation of process errors. We fit 20 different estimation models to each of 100 simulated set of observations with alternative assumptions type and correlation structure of process errors, median natural mortality was known or estimated, and a B-H SRR was assumed or not.

Across simulations we summarized probability of convergence of fitted models, accuracy of marginal AIC in determining the correct process error assumption and it ability to determine the Beverton-Holt SRR, bias in annual spawning stock biomass, in estimation of natural mortality, and in SRR parameters, and severity of retrospective patterns for estimation models.

Alternative measures of convergence performed differently. Invertible hessians and resulting standard error estimation was possible when criteria based on the gradient of the optimized log-likelihood with respect to the fixed effects parameters. Using hessian-based convergence, probability of convergence was best for models that assumed the correct source of process error, assumed M was known, and did not assume SRRs.

Using marginal AIC provided most accurate inferences about the process errors on recruitment and survival and selectivity, in that AIC preferred EMs with assumed process errors that matched OMs, and EMs with those assumed process errors were not preferred when alternative process errors were assumed in the OMs. However, when the true process errors were more variable, AIC accuracy increased to a useful level. We found AIC more accurately determined a B-H SRR rather than the null model without a S-R relationship when there was low variability in recruitment, low variability in survival random effects, and higher variation in spawning biomass over the time series. When the median natural mortality rate was estimated, we found large bias and uncertainty much more likely in model output such as spawning stock biomass.

Bias in spawning stock biomass estimation was generally low for estimation models that assumed the correct source of process error when there was lower observation error. Reliable estimation of SRR parameters only appears possible in ideal situations with lower observation errors in age composition and indices, lower variability in recruitment process errors and large contrast in spawning biomass over time. We found little evidence of bias for many OM process error assumptions when there was contrast in fishing pressure even when there was greater observation error although it can lead to more variable estimates of natural mortality. For OMs where there was bias in natural mortality due to high observation error, estimation the SRR seemed to remove the bias. However, estimation of natural mortality can cause large differences between the true and estimated SSB (that may be unbiased on average) when there is less contrast in fishing pressure over time and higher observation error.

Retrospective patterns were generally weak for all estimation models regardless of the true source of process error, but they can be expected for recruitment even for the correct process error assumptions when observation error is high. When models did exhibit some retrospective pattern, estimation the median natural mortality rate tended to remove it.
-->

@@

This work was funded by NOAA Fisheries Northeast Fisheries Science Center.

<!--
# CRediT authorship contribution statement {-}

**Timothy J. Miller:** Conceptualization, Methodology, Writing - original draft, Formal analysis, Visualization.
 -->

>>>>

@@

We calculated Mohn's $\rho$ for SSB, fully-selected fishing mortality, and recruitment for each EM [@mohn99]. We fit 7 peels for each EM and calculated median 95\% confidence intervals for Mohn's $\rho$ using the same methods as that for relative bias.

@@

@@

@@

This work was funded by NOAA Fisheries Northeast Fisheries Science Center. We thank Jon Deroba for helpful comments on an earlier version of this manuscript.
